Improved state-space models for inference about spatial and temporal variation in abundance from count data

### 2012-09-07 Dail_Mad_ext.tex
ss equations in intro

Theoretical ecology requires models of population dynamics for testing
hypotheses regarding spatial and temporal variation in
abundance. For example, much theoretical work has focused on
understanding the importance and existence of phenomenon such as
density-dependent population regulation, population cycling, and spatial synchrony
\citep{may:1975,royama:1977,turchin:1990,dennis_taper:1994,bjornstad_etal:1999},
and population models are required to evaluate associated hypotheses.
In applied contexts, population models are used for
estimating extinction probabilities
\citep{schoener_spiller:1992,nadeem_lele:2011} and for predicting the
effects of future environmental conditions or conservation actions on
population size \citep{jamieson_brooks:2004,hatfield_etal:2012}.
In order to address these questions, two complicating factors must
be confronted when fitting population models to data.
First, deterministic models of population
dynamics are virtually always inadequate due to process variation,
the inherent stochasticity in demographic parameters and environmental
conditions \citep{bjornstad_grenfell:2001}. %clark_bjornstad:2004}.
Second, abundance---the natural state variable in studies
of population dynamics---can rarely be observed perfectly in field
studies because of observation error, such as imperfect
detection \citep{link_nichols:1994,kery_etal:2009}.
%Failure to account for process
%variation and observation error can bias the estimators of abundance
%and related parameters.

State-space models are a widely used approach for
studying population dynamics while accounting for process variation
and observation error \citep{devalpine_hastings:2002,
  buckland_etal:2004, dennis_etal:2006}. Classical state-space
models are time-series models in which the true state of the
system (e.g. population size during each year) is observed
imperfectly. One reason for the widespread adoption of state-space
models in ecology is that failure to account for process variation and
observation error can bias estimators of abundance and population
growth parameters.
% \citep{kery_etal:2009}.
For instance, the strength of
density dependence will be over-estimated if observation error is
ignored \citep{link_nichols:1994,shenk_etal:1998}.

One of the simplest state-space models can be described as follows.
Let $N_t$ be the abundance of a species during year $t$, for
$t=1,\hdots,T$, and let $X_t$ be
the observed data, which differs from $N_t$ due to observation error,
a random effect denoted $\zeta_t$. Temporal variation in $N_t$ is
modeled using a deterministic population growth model
(e.g. exponential, logistic, etc \dots)~and a stochastic model allowing
for random process variation, which we denote $\grave{o}_t$. As an
example, under exponential population growth, the model could be
written as:
%\begin{gather}
%N_t = N_{t-1}\exp(r)+\grave{o}_t \nonumber \\
%X_t = N_t + \zeta_t
%\label{eq:ss1}
%\end{gather}
\begin{subequations}
  \label{eq:ss1}
  \begin{gather}
    \mu_t = N_{t-1}\exp(r) \label{eq:ss1a} \\ \label{eq:ss1b}
    N_t \sim \text{N}(\mu_t, \grave{o}_t) \\ \label{eq:ss1c}
    X_t \sim \text{N}(N_t, \zeta_t)
  \end{gather}
\end{subequations}
where $r$ is the intrinsic rate of increase. Although the exponential
growth model shown here is often replaced with density dependent
models, most state-space models share the common feature that the
error terms are assumed to be i.i.d random effects distrubted as
$\grave{o}_t \sim \mathrm{N}(0, \sigma)$ and
$\zeta_t \sim \mathrm{N}(0, \tau)$. The Gaussian assumption allows for parameter
estimation using the Kalman filter \citep{dennis_etal:2006}, which is
much more computationally efficient than estimation methods when
random effects are not Gaussian distributed
\citep{devalpine_hastings:2002}.

>>>>

Theoretical ecology requires models of population dynamics for testing
hypotheses regarding spatial and temporal variation in
abundance. For example, much theoretical work has focused on
understanding the importance and existence of phenomenon such as
density-dependent population regulation, population cycling, and spatial synchrony
\citep{may:1975,royama:1977,turchin:1990,dennis_taper:1994,bjornstad_etal:1999},
and population models are required to evaluate associated hypotheses.
In applied contexts, population models are used for
estimating extinction probabilities
\citep{schoener_spiller:1992,nadeem_lele:2011} and for predicting the
effects of future environmental conditions or conservation actions on
population size \citep{jamieson_brooks:2004,hatfield_etal:2012}.
In order to address these questions, two complicating factors must
be confronted when fitting population models to data.
First, deterministic models of population
dynamics are virtually always inadequate due to process variation,
the inherent stochasticity in demographic parameters and environmental
conditions \citep{bjornstad_grenfell:2001}. %clark_bjornstad:2004}.
Second, abundance---the natural state variable in studies
of population dynamics---can rarely be observed perfectly in field
studies because of observation error, such as imperfect
detection \citep{link_nichols:1994,kery_etal:2009}.

State-space models are a widely used approach for
studying population dynamics while accounting for process variation
and observation error \citep{devalpine_hastings:2002,
  buckland_etal:2004, dennis_etal:2006}. Classical state-space
models are time-series models in which the true state of the
system (e.g. population size during each year) is observed
imperfectly. One reason for the widespread adoption of state-space
models in ecology is that failure to account for process variation and
observation error can bias estimators of abundance and population
growth parameters. For instance, the strength of
density dependence will be over-estimated if observation error is
ignored \citep{link_nichols:1994,shenk_etal:1998}.

One of the simplest state-space models can be described as follows.
Let $N_t$ be the abundance of a species during year $t$, for
$t=1,\hdots,T$, and let $X_t$ be
the observed data, which differs from $N_t$ due to observation error,
a random effect denoted $\zeta_t$. Temporal variation in $N_t$ is
modeled using a deterministic population growth model
(e.g. exponential, logistic, etc \dots)~and a stochastic model allowing
for random process variation, which we denote $\grave{o}_t$. As an
example, under density-independent exponential population growth, the
model could be written as:
\begin{equation}
%  \label{eq:ss1}
  \left.
  \begin{aligned}
    \mu_t &= N_{t-1}\exp(r) \label{eq:ss1a} \\ %\label{eq:ss1b}
    N_t &\sim \text{N}(\mu_t, \grave{o}_t) \\ %\label{eq:ss1c}
    X_t &\sim \text{N}(N_t, \zeta_t)
  \end{aligned} \;
  \right\} \; \text{for} \; t=2,\hdots,T
\end{equation}
%\begin{subequations}
%  \label{eq:ss1}
%  \left.
%  \begin{gather}
%    \mu_t = N_{t-1}\exp(r) \label{eq:ss1a} \\ \label{eq:ss1b}
%    N_t \sim \text{Normal}(\mu_t, \grave{o}_t) \\ \label{eq:ss1c}
%    X_t \sim \text{Normal}(N_t, \zeta_t)
%  \end{gather}
%  \right\} \text{for}
%\end{subequations}
where $r$ is the intrinsic rate of increase. Note that no model is
presented for the initial state at $t=1$---this issue will be
discussed subsequently. Note also that the exponential
growth model shown in Eq.~\ref{eq:ss1a} can be easily replaced with
any population growth model, including density-depedent models. This
flexibility is an appealing feature of state-space models. One feature
that is common to classical state-space models is that the
error terms are assumed to be i.i.d random effects following Gaussian
(Normal) distributions:
$\grave{o}_t \sim \mathrm{N}(0, \sigma)$ and
$\zeta_t \sim \mathrm{N}(0, \tau)$. The Gaussian assumption allows for parameter
estimation using the Kalman filter \citep{dennis_etal:2006}, which is
much more computationally efficient than estimation methods when
random effects are not Gaussian distributed
\citep{devalpine_hastings:2002}.
